---
title: |
  |  \Large A Comparison of Batter Performance and League Quality Across Domestic T-20 Cricket Leagues
author: | 
  | \large Matthew Stuart$^{1,2}$, Hassan Raffique$^{3}$, Leigha DeRango$^{1,2}$
  | \large Gregory J. Matthews$^{1,2}$
  | \vspace{-1.1mm}
  | \large $^1$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Syracuse University, Syracuse, NY, USA \vspace{-1.1mm}
  | \large $^+$ Corresponding: mstuart1@luc.edu \vspace{-1.1mm}
abstract: |
  | Wicked Googly \vspace{2mm}
  | *Keywords*: Cricket
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \usepackage{amssymb}
 \usepackage{amsmath}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

\newcommand{\iid}{\overset{iid}{\sim}}

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
```

Abstract:

This paper investigates batter performance across major domestic Twenty20 (T20) cricket leagues to understand how player effectiveness varies by competition context. Using a comprehensive dataset comprising individual match-level batting records from the Indian Premier League (IPL), Big Bash League (BBL), Caribbean Premier League (CPL), Pakistani Super League (PSL), and South Africa 20 (SAT), we fit a multinomial regression model for the number of runs scored per bowled ball accounting for batter and league conditions as well as in-match effects. A hierarchical modeling framework utilizing a Bayesian framework is employed to quantify both player-specific ability and league-specific correlations, allowing for quantifiable player comparisons across leagues and the estimation of league-adjusted performance ratings. Our findings will offer insights into talent evaluation and the transferability of performance across T20 competitions.

\newpage

# Introduction {#sec:intro}
openWAR and cricWAR.  

In the game of cricket, the number of runs scored on a particular pitch typically ranges between 0 and 6 (though theoretically values larger than 6 are possible, they are rare and do not occur at all in our particular data set).  

# Data {#sec:data}


```{r echo = FALSE}
bbb <- read.csv("../data/cricket_data.csv")
bbb <- bbb %>% 
  filter(year >= 2015) %>%
  arrange(year)
library(tidyverse)

#table(ipl$season)
#Are the teams  from Delhi and Pujarat separate teams?  Or same team different name?  
#ipl %>% group_by(batting_team, season) %>% summarise(n = n()) %>% group_by(batting_team) %>% summarise(n = n())
#ipl %>% group_by(batting_team, season) %>% summarise(n = n()) %>% group_by(season) %>% summarise(n = n()) 
```

```{r}
bbb %>% group_by(batting_team, year, league) %>% summarise(n = n()) %>% group_by(year,league) %>% summarise(n_teams = n()) %>% pivot_wider(names_from = league,values_from=n_teams,values_fill=NA)
```

We have data from the Indian Premier League (IPL),  Big Bash League (BBL), Caribbean Premier League (CPL), Pakistani Super League (PSL), and South Africa 20 (SAT), consisting of `r nrow(bbb)` balls thrown, providing $L = 5$ leagues worth of data. From 2015 - 2021, the IPL, the longest running T20 league, had 8 teams followed by 10 teams in the 2022 season.  In addition, the SAT was first contested in the 2022-2023 season.
```{r}
bbb %>% group_by(league,year) %>% summarize(n_pitches = n()) %>% pivot_wider(names_from = league,values_from = n_pitches)
```
Runs in cricket are either scored by running back and forth between the wickets once the ball is put into play (generally resulting in 1 or 2 runs, but theoretically any value is possible).  In addition, a ball that is hit in the air over the boundary (termed a "boundary") is worth 6 runs and if the ball rolls to the boundary or bounces in the field of play and then clears the boundary this is worth 4 runs (termed a "boundary 4").  As a result the distribution of runs scored on a a particular pitch has large peaks are 0 and 1 with a big drop off from 1 to 2.  Values of 3 and 5 are extremely rare accounting for only `r round(mean(bbb$runs_off_bat == 3)*100,2)`% and `r round(mean(bbb$runs_off_bat == 5)*100,2)`% of values across all pitches in our data set.  Values of 4 and 6 spike because of boundaries and boundary fours and together account for `r round(mean(bbb$runs_off_bat == 4)*100 + mean(bbb$runs_off_bat == 6)*100,2)`% of all values.  

```{r bar, fig.cap = "Bar Plot of the number of runs scored of a particular ball"}
bbb %>% ggplot(aes(x = runs_off_bat)) + geom_bar()
```
<!-- Figure \ref{fig:bar} shows a bar plot.   -->

```{r eval=FALSE, fig.cap="Can I caption a table?"}
bbb %>% group_by(striker) %>% summarize(xbar = mean(runs_off_bat), n = n()) %>% arrange(-xbar) %>% filter(n > 500)
```

# Models {#sec:models}

Figure \ref{fig:bar} displays a histogram of the number of runs scored per ball in IPL, BBL, CPL, PSL, and SAT matches from 2015-2024, consisting of $n = 408,056$ balls thrown. It is likely that a distribution used for modelling counts, such as the Poisson distribution, will violate the necessary assumptions. For this reason, we treat this as a classification problem and fit the number of runs scored per ball, $Y_i$, by a multinomial distribution.  In addition, we exclude any balls that scored three or five runs because of their prementioned rarity of occuring. Thus, we are left with a dataset consisting of $n = 406,454$ balls thrown.

We utilize a mixed effects model, incorporating fixed effects for the general in-match situations as well as random effects for the variability of the bowler, batter, and runner. Denote $Y_i$ as the number of runs scored on ball $i = 1,\dots,n$ and $\boldsymbol{X}_i$ as the vector of covariates for the fixed effects of ball $i$. Table \@ref(tab:covariates) provides a description of the covariates for the fixed effects of our model. The model is specified with four logit transformations relative to the event $Y_i = 0$, or, written explicitly
\begin{equation}
\log\left(\frac{P(Y_i = k | \boldsymbol{X}_i)}{P(Y_i = 0 | \boldsymbol{X}_i)}\right) = \boldsymbol{X}_i \boldsymbol{\beta}_k + u_{k, b_it_i} + s_{k,v_i} \label{model}
\end{equation}
for $k \in \{1,2,4,6\}$ where $\boldsymbol{\beta}_k$ is the fixed effect for scoring $k$ runs, $u_{k,b_it_i}$, is the random effect for the batter($b_i$) and year($t_i$), and $s_{k,v_i}$ is the random effect for venue for ball $i$, respectively.  For the batter random effects ${u}_{k,bt}$, we model the random effects for each $b$ via an AR1 process.  More specifically, for each $k \in \{1,2,4,6\}$, we set
\begin{align}
{u}_{k,bt} & \sim \mathcal{N}(\rho_k {u}_{k,b(t-1)},\sigma_{k}), \nonumber \\ 
{u}_{k,b1} & \sim \mathcal{N}(0,\frac{\sigma_{k}}{1-\rho_k^2})\label{ranefs_bat}
\end{align}

for $t = 2,\cdots,T$. For the random effects for venue, we assume
\begin{align}
{s}_{k,v} & \sim \mathcal{N}(0,\tau_{k}), \label{ranefs_ven}
\end{align}

Given the size of the random effects, the size of our free, unknown parameter vector $\boldsymbol{\Theta} = \{\boldsymbol{\beta}_k,\boldsymbol{u}_k, \boldsymbol{v}_k,\rho_k,\sigma_k,\tau_k: k \in \{1,2,4,6\}\}$ is 43,740. To handle such a large parameter vector as well as the complicated structure of our model, we perform a Bayesian analysis on the data with prior distributions and sampling procedure outlined in the supplemental file.<!-- However, not all batter-league-season combinations are observed. To reduce the parametric burden, we only estimate the posterior of $\boldsymbol{u}_k$ for each batter in the years in which they bat. Therefore, in practice, we have 119,164 parameters, reducing our parameteric burden by over 66%.-->  

```{r tab:covariates, results='asis'}
library(knitr)
data.frame(Variable = c("First Innings",
                        "Balls Remaining",
                        "Runs to Win",
                        "Runs Scored",
                        "Wickets Lost",
                        "League"),
           `Variable Description` = c("Indicator for the 1st innings of the match",
                                      "Number of balls remaining in the innings",
                                      "Number of runs remaining to score to win the match (2nd innings)",
                                      "Number of runs scored in the innings up to current ball",
                                      "Number of wickets lost in the innings up to current ball",
                                      "League in which the match is played")) %>% kable(caption = "Description of covariates for fixed effects of model")
```

<!-- Let $r_i$ be the number of runs scored on the $i$-th ball with $i=1,\cdots,n$ and $r_i \in \left\{0,1,\cdots,6\right\}$.  Then define ${\bf y_{i}} = \left(y_{i0},\ldots,y_{i6}\right)$ where $y_{ij} = I(r_i = j), \forall j = 0, \cdots, 6$ and $I(.)$ is the indicator function.  We then model:  -->

<!-- $$ -->
<!-- {\bf y_i} \sim MN(1,{\bf p_i}) -->
<!-- $$ -->
<!-- where ${\bf p_{i}} = \left(p_{i0},\ldots,p_{i6}\right)$, $p_{ij}$ is the probability that $j$ runes are scored on the $i$-th ball, and $\sum_{j = 0}^{6} p_{ij} = 1$ for all $i$.   -->

<!-- $$ -->
<!-- log\left(\frac{p_{j}}{p_{0}}\right) = \beta_{0j} + {\bf X}{\bf \beta}_j + b_{j,batter} + b_{j,bowler} + b_{j,runner} -->
<!-- $$ -->
<!-- for $j = 1,\ldots,6$ where $\beta_{0j}$ is the intercept of the $j$-th linear component, ${\bf \beta}_j$ is a $P$-dimensional vector containing the regression coefficients for the fixed effects, ${\bf X}$ is the matrix of covariates, and $b_{j,batter}, b_{j,bowler}, b_{j,runner}$ are random effects for the batter, bowler, and runner, respectively. -->

<!-- ## Priors -->
<!-- whatever priors we decide to use-->



# Results {#sec:results}

# Discussion, Future work and conclusions {#sec:conclusions}

# Acknowledgements {-}

# Supplementary Material {-}

All code for reproducing the analyses in this paper is publicly available at https://github.com/gjm112/cricketIPL

## Bayesian priors and posterior sampling

In the data analysis outlined in Section \ref{sec:models} of the main manuscript, we set the following prior distributions:
\begin{align}
  \beta_{p,k} & \iid \mathcal{N}(0,5),\nonumber\\
  \sigma_{k} & \iid \mathcal{N}(0,1)I(\sigma_{k} > 0),\nonumber\\
  \tau_{k} & \iid \mathcal{N}(0,1)I(\tau_{k} > 0),\nonumber\\
  \rho_{k} & \iid \mathcal{N}(1,1)I(-1 < \rho_{k} < 1),
\end{align}
for $p = 1,\dots,10$ and $k \in \{1,2,4,6\}$. We used STAN to obtain 4000 posterior draws from our parameter set $\boldsymbol{\Theta} = \{\boldsymbol{\beta}_k,\boldsymbol{u}_k, \boldsymbol{v}_k,\rho_k,\sigma_k,\tau_k: k \in \{1,2,4,6\}\}$
<!-- , where LKJ is the Lewandowski-Kurowicka-Joe distribution for correlation matrices (@LKJ). We update the model parameters $\{\boldsymbol{\beta}_k,\boldsymbol{\tau}_k,R_k,\Phi_k: k \in \{1,2,4,6\}\}$ individually using a Metropolis-adjusted Langevin algorithm (MALA).  MALA is a version of a Metropolis Hastings algorithm where the new states are proposed using overdamped Langevin dynamics.  More specifically, at step $r$ of the algorithm, for each parameter $\boldsymbol{\Theta}$, we sample a proposal $$\boldsymbol{\Theta}^* \sim \mathcal{N}\left(\boldsymbol{\Theta}_r + a\nabla \log \pi(\boldsymbol{\Theta}_r|\boldsymbol{y},\boldsymbol{X}), \sqrt{2a}\right)$$ where $\pi$ is the functional form of the posterior distribution for $\boldsymbol{\Theta}$ and $a$ is a tuning parameter for the proposal distribution.  The tuning paramter $a$ is chosen via an adaptation of the primal-dual algorithm from @Nesterov2009, which was also utilized in @NUTS, to obtain an acceptance probability of 0.574.  For the random effects $\boldsymbol{u}$, we use a Particle Gibbs with ancestor sampling (PGAS) algorithm (@PGAS), which is a variation of sequential Monte Carlo (SMC) algorithms that enables fast mixing of time series data. For the PGAS, we incorporate a PGAS particle size $N = 10$. -->



# References
