---
title: "data cleaning"
author: "Leigha DeRango"
date: "2023-07-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# libraries
```{r}
library(dplyr)
library(cricketdata)
```


# get leagues individually bbb
```{r}
# indian premier league
ipl_bbb = fetch_cricsheet('bbb', 'male', 'ipl') %>%
  mutate(league = 'IPL')

# big bash league australia
bbl_bbb = fetch_cricsheet('bbb', 'male', 'bbl') %>%
  mutate(league = 'BBL')

# pakistan super league
psl_bbb = fetch_cricsheet('bbb', 'male', 'psl') %>%
  mutate(league = 'PSL')

# caribbean premier league
cpl_bbb = fetch_cricsheet('bbb', 'male', 'cpl') %>%
  mutate(league = 'CPL')

# south african league
sat_bbb = fetch_cricsheet('bbb', 'male', 'sat')%>%
  mutate(league = 'SAT')


bbb = rbind(ipl_bbb, bbl_bbb, psl_bbb, cpl_bbb, sat_bbb)
names(bbb)
```

# lists of players per match per season
```{r}
### IPL
ipl_players = fetch_cricsheet('player', 'male', 'ipl')
ipl_matches = fetch_cricsheet('match', 'male', 'ipl') %>% select(match_id, season)

ipl_full = left_join(ipl_players, ipl_matches, by = 'match_id', keep = TRUE) %>% select(team, player, match_id.x, season) %>% mutate(season_adj = substr(season, 1, 4), league = 'IPL')

### BBL
bbl_players = fetch_cricsheet('player', 'male', 'bbl')
bbl_matches = fetch_cricsheet('match', 'male', 'bbl') %>% select(match_id, season)

bbl_full = left_join(bbl_players, bbl_matches, by = 'match_id', keep = TRUE) %>% select(team, player, match_id.x, season) %>% mutate(season_adj = substr(season, 1, 4), league = 'BBL')

### PSL
psl_players = fetch_cricsheet('player', 'male', 'psl')
psl_matches = fetch_cricsheet('match', 'male', 'psl') %>% select(match_id, season)

psl_full = left_join(psl_players, psl_matches, by = 'match_id', keep = TRUE) %>% select(team, player, match_id.x, season) %>% mutate(season_adj = substr(season, 1, 4), league = 'PSL')

### CPL
cpl_players = fetch_cricsheet('player', 'male', 'cpl')
cpl_matches = fetch_cricsheet('match', 'male', 'cpl') %>% select(match_id, season)

cpl_full = left_join(cpl_players, cpl_matches, by = 'match_id', keep = TRUE) %>% select(team, player, match_id.x, season) %>% mutate(season_adj = substr(season, 1, 4), league = 'CPL')

### SAT
sat_players = fetch_cricsheet('player', 'male', 'sat')
sat_matches = fetch_cricsheet('match', 'male', 'sat') %>% select(match_id, season)

sat_full = left_join(sat_players, sat_matches, by = 'match_id', keep = TRUE) %>% select(team, player, match_id.x, season) %>% mutate(season_adj = substr(season, 1, 4), league = 'SAT')

# players of each match by season
combined_full = rbind(ipl_full, bbl_full, psl_full, cpl_full, sat_full)
  # team/player/match_id/season/season_adj/league
```

# players total games in league by season
```{r}

games_per_season = combined_full %>%
  group_by(season_adj, player) %>%
  mutate(games_in_season = n()) %>% 
  ungroup() %>% 
  group_by(season_adj, player, league) %>% 
  mutate(games_in_league = n()) %>%
  select(player, season_adj, league, games_in_league, games_in_season) %>%
  distinct() %>%
  arrange(player, season_adj)

  # player/season_adj/league/games_in_league/games_in_season

```

# only players that play in more than one league per season
```{r}

many_leagues = games_per_season %>% filter(games_in_league != games_in_season)
  # player/season_adj/league/games_in_league/games_in_season

```

# add position and ids to those that overlap in leagues 
```{r}

positions = left_join(many_leagues, player_meta, by = join_by(player == unique_name)) %>%
  select(player, season_adj, league, games_in_league, games_in_season, cricinfo_id,
         cricsheet_id, batting_style, bowling_style, playing_role)

```

# batter/pitcher matchup counts by games and pitches faced per season
```{r}

ipl_summary = ipl_bbb %>% mutate(season_adj = substr(season, 1, 4)) %>% 
  group_by(bowler, striker, season_adj) %>%
  mutate(pitches_in_league = n(), .keep = 'all') %>% 
  ungroup() %>% group_by(match_id) %>%
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>% ungroup() %>%
  group_by(bowler, striker, season_adj) %>% 
  mutate( games_in_league = n(), .keep = 'all') %>%
  mutate(league = 'IPL', .before = games_in_league) %>% 
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>%
  select(season_adj, bowler, striker, league, games_in_league, pitches_in_league)
  
bbl_summary = bbl_bbb %>% mutate(season_adj = substr(season, 1, 4)) %>% 
  group_by(bowler, striker, season_adj) %>%
  mutate(pitches_in_league = n(), .keep = 'all') %>% 
  ungroup() %>% group_by(match_id) %>%
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>% ungroup() %>%
  group_by(bowler, striker, season_adj) %>%
  mutate( games_in_league = n(), .keep = 'all') %>%
  mutate(league = 'BBL', .before = games_in_league) %>% 
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>%
  select(season_adj, bowler, striker, league, games_in_league, pitches_in_league)
  
psl_summary = psl_bbb %>% mutate(season_adj = substr(season, 1, 4)) %>% 
  group_by(bowler, striker, season_adj) %>%
  mutate(pitches_in_league = n(), .keep = 'all') %>% 
  ungroup() %>% group_by(match_id) %>%
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>% ungroup() %>%
  group_by(bowler, striker, season_adj) %>%
  mutate( games_in_league = n(), .keep = 'all') %>%
  mutate(league = 'PSL', .before = games_in_league) %>% 
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>%
  select(season_adj, bowler, striker, league, games_in_league, pitches_in_league)
  
cpl_summary = cpl_bbb %>% mutate(season_adj = substr(season, 1, 4)) %>% 
  group_by(bowler, striker, season_adj) %>% 
  mutate(pitches_in_league = n(), .keep = 'all') %>% 
  ungroup() %>% group_by(match_id) %>%
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>% ungroup() %>%
  group_by(bowler, striker, season_adj) %>%
  mutate( games_in_league = n(), .keep = 'all') %>%
  mutate(league = 'CPL', .before = games_in_league) %>% 
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>%
  select(season_adj, bowler, striker, league, games_in_league, pitches_in_league)
  
sat_summary = sat_bbb %>% mutate(season_adj = substr(season, 1, 4)) %>% 
  group_by(bowler, striker, season_adj) %>% 
  mutate(pitches_in_league = n(), .keep = 'all') %>% 
  ungroup() %>% group_by(match_id) %>%
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>% ungroup() %>%
  group_by(bowler, striker, season_adj) %>%
  mutate( games_in_league = n(), .keep = 'all') %>%
  mutate(league = 'SAT', .before = games_in_league) %>% 
  distinct(pick('bowler', 'striker'), .keep_all = TRUE) %>%
  select(season_adj, bowler, striker, league, games_in_league, pitches_in_league)

#format: 
# group by bowler, striker, season and count number of rows for pitches faced, then ungroup

# group by match_id and pick one row for each distinct bowler striker matchup in that game, then ungroup

# group by bowler, striker, and season and count rows for the number of games faced

# add league variable


lifetime_matchups = rbind(ipl_summary, bbl_summary, psl_summary, cpl_summary, sat_summary)
  #season/batter/pitcher/league/games faced in league/pitches faced in league
```


# count of batter/pitcher matchups across leagues by season
```{r}

league_matchups = lifetime_matchups %>%
  mutate(IPL = ifelse(league == 'IPL', 1, 0),
         BBL = ifelse(league == 'BBL', 1, 0),
         PSL = ifelse(league == 'PSL', 1, 0),
         CPL = ifelse(league == 'CPL', 1, 0),
         SAT = ifelse(league == 'SAT', 1, 0)) %>%
  group_by(bowler, striker, season_adj) %>%
  summarize(IPL_matchup = sum(IPL),
            BBL_matchup = sum(BBL),
            PSL_matchup = sum(PSL), 
            CPL_matchup = sum(CPL), 
            SAT_matchup = sum(SAT))
  #batter/pitcher/IPL matchups/BBL matchups/PSL matchups/...

```

# analysis of batter/pitcher matchup overlap between leagues
```{r}

sum_leagues = league_matchups %>% 
  mutate(total_leagues = IPL_matchup + BBL_matchup + PSL_matchup + CPL_matchup + SAT_matchup)

hist(sum_leagues$total_leagues)
summary(sum_leagues$total_leagues)

#average number of leagues a player was in by season
season_averages = sum_leagues %>%
  group_by(season_adj) %>%
  summarize(average_leagues = mean(total_leagues)) %>%
  print()
  # 2016 had the most overlap
  
```









